### 2019-09-30 submission/manuscript.tex
Editing
Edited manuscript for wording, structure, and to add references.

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

@@

\hypertarget{observation-format}{%
\subsection{Observation format}\label{observation-format}}

@@

75 words

@@

3000 words total

Few scientists set out on a nearly two decade long journey with a
specific goal in mind. Often we fail to start a scientific journey
because it looks too hard. Perhaps we get bogged down in all of the
things that could go wrong. Perhaps we go astray from the path because
we find something else that appears more interesting. Every scientist
picks their own path and takes their own forks in the road. From the
outside, it may appear to be a random walk. Nevertheless, these
meandering journeys in are common in science.

At the risk of navel gazing, looking back on our scientific journeys can
be instructive to other scientists who are overwhelmed at the prospect
of looking forward at their careers {[}Lenski and Chemistry guy{]}. By
no means is my personal journey over, but since 2002 I have been on a
journey that I did not realize I was on. Now that the paper introducing
the mothur software package is ten years old and has become the most
cited paper published by \emph{Applied and Environmental Microbiology},
it is worth stepping back and using the development of mothur as a story
that likely has parallels to many other research stories that have taken
time to develop.

I fondly recall preparing a poster for the 2002 meeting of research
groups supported by the NSF-supported Microbial Observatories Program. I
wanted to triumphantly show that I had sequenced more than 600 16S rRNA
gene sequences from a single 0.5-g sample of Alaskan soil. This was
greater sequencing depth than anyone else had achieved for a single
sample. As I was preparing the poster, I walked into the office of Jo
Handelsman, my postdoctoral research advisor, and laid out the outline
for the poster. She asked if I could add one of those ``curvy things'',
a rarefaction curve, to show where I was in sampling the community.
Rarefaction curves and attempts to estimate the taxonomic richness of
soil had become popular because of the simple, but impactful mini-review
by Jennifer Hughes and her colleagues, which introduced the field to
operational taxonomic units (OTUs), rarefaction curves, and richness
estimates {[}DOI: 10.1128/AEM.67.10.4399-4406.2001{]}. I do not recall
whether that poster had a rarefaction curve on it, but her question
primed my career.

\textbf{\emph{Introducing DOTUR and friends.}} When Jo asked me to
generate a rarefaction curve for the poster, the request was not
trivial. How would I bin the sequences into OTUs? Hughes and her
colleagues did it manually for datasets that had fewer than 284
sequences. Although I could possibly do that for my 600 sequences, my
goal was to generate 1,000 sequences from the sample and to repeat that
sampling effort for other samples. I needed something that could be
automated. Furthermore, the software that Hughes used, EstimateS,
required a series of tedious data formatting steps. I had found my first
problem. How would I assign sequences to OTUs and use that data to
estimate the richness and diversity of a sample? The second problem
would be how could I compare the sequences found in one sample to
another sample? The solution to the first problem, DOTUR (Distance-based
OTUs and Richness), took us two years to develop
{[}10.1128/AEM.71.3.1501-1506.2005{]}. DOTUR did two things: given a
matrix describing the genetic distance between pairs of sequences, it
would cluster those sequences into OTUs for any distance threshold to
define the OTUs and then it would use the frequency of each OTU to
calculate a variety of alpha diversity metrics. The solutions to the
second problem would come from our work to develop software including
S-LIBSHUFF {[}2004{]}, SONS (Shared OTUs and Similarity) {[}2006{]}, and
TreeClimber {[}2006{]}. Around the same time, Catherine Lozupone and Rob
Knight were developing their UniFrac tools to compare communities with a
phylogenetic rather than OTU-based approach {[}PMID: 16332807; PMID:
17220268{]}. With these tools, the field of microbial ecology had a
quantitative toolbox for describing and comparing microbial communities.

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification {[}XXXX{]}. We knew there were chimeras {[}XXXX{]}.
Getting to the distance matrix required trimming and correcting sequence
errors, aligning them, and finding the most appropriate way to calculate
a pairwise distance. It was also rare to have experimental replication
to perform statistical tests to compare treatment groups. We frequently
used a dataset comparing Scottish soils from from Alison McCaig and
colleagues. This dataset consisted of two experimental groups, each
replicated three times with 45 sequences in each replicate.
Nevertheless, we had excuses and work arounds for these problems that
served our needs. At the time, I felt that the biggest problems were how
to cluster the sequences into OTUs and how to use those clusterings to
test our hypotheses. Along the way we would demonstrate the utility of
such tools to answer questions like where are we in the bacterial
census? How many sequences would it take to see every OTU in that sample
of Alaskan soil? How does the word usage of \emph{Goodnight, Moon}
compare to that of \emph{Portrait of a Lady}? More importantly, 1,900
papers used DOTUR to facilitate their own research questions. Had we
waited to solve all of the problems that plague 16S rRNA gene
sequencing, we would still be waiting.

As we developed these tools, I found a unique niche in microbiology. I
believe that my undergraduate and graduate training as a biological
engineer prepared me to think about research questions from a systems
perspective, to think quantitatively, and to understand the value of
using computer programs to help solve problems. As an undergraduate
engineering student, I learned the Pascal programming language and
promptly forgot much of it as an engineering graduate student. As a
postdoc, I learned the Perl programming language to better understand
how LIBSHUFF, a tool for comparing the structure of two communities,
worked since it was written in Perl {[}DOI:
10.1128/AEM.70.9.5485-5492.2004{]}. After writing my own version of
LIBSHUFF and seeing the speed of the version written in C++ by my
collaborator, Bret Larget, I converted my Perl version of DOTUR into
C++. At the time, the conversion from Perl to C++ seemed like an
academic exercise to learn a new language. My Perl version only took a
minute or so to process the final collection of 1,000 sequences and the
C++ version took seconds. Was that really such a big difference? In
hindsight, as we now process datasets with millions of sequences, the
decision to learn to C++ was critical. The ability to pick up computer
languages to solve problems was enabled by my prior training. It was
also a skill that was virtually unheard of in microbiology. Today
researchers without the ability to program in Python or R are at a
significant disadvantage.

\textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
received an email from Mitch Sogin asking whether DOTUR could handle
more than a million sequences. Without answering his question, I asked
where he found a million sequences. Little did I know that his email
would represent another pivot in the development of these tools. His
group would be the first to use 454 sequencing technology to generate
16S rRNA gene sequences {[}PMID: 16880384{]}. Although mothur could
assign those sequences to OTUs, it was slow and required a significant
amount of RAM. As I left my postdoc to start my independent career
across the state from Sogin's lab at the University of Massachusetts in
Amherst, my plan was to rewrite DOTUR, SONS, S-LIBSHUFF, and TreeClimber
for the new world of massively parallelized sequencing. The new tool was
mothur.

Milling about at a poster session at an ASM General Meeting in New
Orleans, I again ran into Mitch who asked what my plans were for new
tools. I told him that I wanted to make a tool like ARB (a powerful
database tool and phylogenetics package), but for microbial ecology
analysis. His retort was, ``You and what army?'' To that point, I had
written every line of code and been answering many emails from people
asking for help. It would be difficulty, but I needed to learn to let go
and share the development process with someone else. He was right, I
would need an army. That ``army'' ended up being Sarah Westcott who has
worked on the mothur project largely from its inception. Today, mothur
is over 200,000 lines of code and Sarah has touched or written nearly
every line of code. Beyond writing and testing mothur's code base, she
has become a conduit for many learning the tools of microbial ecology by
patiently answering questions via email and the package's discussion
forum. The community and I are lucky that Sarah has stayed with the
project for more than a decade. To be honest, such dependency on a
single person makes the project brittle. In hindsight, it would have
been better to have developed mothur with more of an ``army'' or team so
that there is overlap in people's understanding of how mothur works.
Although such an arrangement might work in a software engineering firm,
it is not practical in an academic setting where funding is limited for
developers of free, open source software packages. There are certainly
projects that make this work, but they are rare.

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to third parties
to contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists with any programming skills, expecting that
community to provide contributors that can write code in a syntax that
prizes execution efficiency over developer efficiency was not likely. In
contrast, the QIIME development team could be more distributed because
their code base was primarily written in Python, which prizes developer
efficiency over execution efficiency and exists as a series of wrappers
to execute other developers' code. These choices resulted in many
tradeoffs that have impacted ease of installation, usability, execution
speed, and flexibility. If we were offered a grant to rewrite mothur, we
would likely rewrite it as an R package that leaned heavily on the R
language's C++ packages. Of course, such choices are always best in
hindsight and when we started developing mothur, the ability to
interface between scripting languages like R and Python and C++ code was
not as well developed as it is today, For example, the modern version of
the Rcpp package was first released in 2009 and its popularity was not
immediate. Again, the development of mothur has been a product of the
environment that it was created in. Although these decisions have
largely had positive outcomes, there have been tradeoffs that caused us
to sacrifice other goals.

Beyond contributing to the mothur code base, we sought out other ways to
include the community as developers. The paper describing mothur
included \textbf{N} co-authors, all but three (Schloss, Ryabin, and
Westcott) responded to a call to provide a wiki page that described how
they used an early version of mothur to analyze a data set. Our vision
was that authors might use the mothur wiki to document reproducible
workflows for papers using mothur but to also provide instructional
materials for other seeking to adapt mothur for their uses. Again, this
vision was a product of the environment. IPython notebooks (2011) and R
markdown (2012) would not be developed until later. Unfortunately, once
the incentive of co-authorship was removed, researchers stopped
contributing their workflows to the wiki. Part of the difficulty of
recruiting wiki contributors was a perception by some that the wiki was
not a community resource. For example, I would frequently receive emails
from people telling me that there was a typo on a specific page when the
intention was that they could correct the typos without my input. We
have been more successful in soliciting input and contributions from the
user community through the mothur discussion forum and GitHub issue
tracker. As mothur has matured, we have been dependent on the user
community to use these resources to tell us what features they would
like to see included in mothur. The user community also tells us where
our documentation is confusing. Often we can count on people not
directly affiliated with mothur to provide instruction and their own
experience to other users. We are constantly trying to recruit our
``army'' and are happy to take any contributions we can. Whether the
contributions are to the code base, discussion forum, or suggestions for
new tools, these contributions have been invaluable to the growth and
popularity of mothur.

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and venn diagrams depicting the
overlap between microbial communities. I quickly realized that I would
never put a mothur-generated figure into a manuscript I wrote. Such
visuals require far too much customization to be publication-quality.
Although QIIME has incorporated visualization tools through the Emperor
package, the challenge of users taking default values has downsides as
ordinations with black background or publishing 3-D ordinations in a 2-D
medium litter the literature. Instead, we have encouraged users to use R
packages to visualize mothur-generated results using the minimalR
instructional materials. A second experiment we attempted was to create
a graphical user interface (GUI) for running mothur. Forcing users to
interact with mothur through the command line has been a significant
hurdle for many. Unfortunately, the development effort required to
create and maintain a GUI is significant and there is limited funding
for such efforts. The newest version of QIIME (version 2) has emphasized
interaction with the tools through a GUI and it remains to be seen how
this experiment will go. Another downside of using a GUI is that there
is a risk that reproducibility will suffer if users do not have a
mechanism to document their mouse clicks. This documentation is explicit
in mothur as all commands and output is recorded in a logfile. Given the
heightened recent focus on reproducibility we have extended significant
effort in developing instructional materials teaching users how to
organize, document, and execute reproducible pipelines that allow a user
to go from raw sequence data to a compiled manuscript with figures
through the Riffomonas project. Finally, we collaborated with
programmers through Google Summer of Code to develop commands in mothur
to implement the random forest and SVM machine learning algorithms.
Similar to the challenges of developing attractive visuals, fitting the
algorithms' hyperparameters, testing, and deploying the resulting models
requires a significant amount of customization. Furthermore, this is an
active area of research where methods are still being developed and
improved. Thankfully, there are numerous R and Python packages that do a
better job of developing these models. Again, we have put our efforts
into developing instructional materials that mothur users can use to fit
such models to their data. In each of our ``failed'' experiments, the
real problems were straying from what mothur does well and failing to
grasp what we really wanted the innovation to do. In hindsight, our
solution to these failure has been to provide tutorials to as a conduit
between mothur and their goals.

\textbf{\emph{Competition is good and healthy.}}

\begin{itemize}
\tightlist
\item
  A lot of people have been working in this space - e.g.~clustering
  algorithms
\item
  Approaches to developing software have differed (e.g.~web-based
  (greengenes/RDP), boutique (Quince), wrappers (QIIME))
\end{itemize}

What are we proud of? * Open source packages * Platform independent *
Development of open instructional materials * Data driven development of
methods * Database independent - allows people outside of our field to
use mothur * Reproducibility * Helped create standards - when something
is this popular, you force people to use a certain suite of tools and
methods

The future * 16S rRNA gene sequencing is not dead * Always improving
sequencing methods bring new problems * Larger datasets, need a
replacement for naive open and closed reference clustering * Lingering
controversies - singletons, ASVs, contamination

Longterm funding * The development of mothur was enabled by initial
funding from a grant to Sogin from the Sloan Foundation to support his
VAMPS (Visualization and Analysis of Microbial Population Structures)
initiative. * NSF * NIH * Bootstrapping from other projects

Red Queen

>>>>

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font

%\usepackage{mathpazo} % Palatino font
%\renewcommand*\familydefault{\rmdefault} % Use the roman version of the font

\usepackage[T1]{fontenc}

@@

\hypertarget{mini-review}{%
\subsubsection{Mini-review}\label{mini-review}}

@@

250 words

@@

Few scientists set out on a nearly two decade-long journey with a
specific goal in mind. Often we fail to start a scientific journey
because it looks too hard. Perhaps we get discouraged by all of the
things that could go wrong. Maybe we stray from the path because we find
something that is more interesting. Every scientist picks their own path
and takes their own forks in the road. From the outside, it may appear
to be a random walk. Nevertheless, these meanderings are common in
science.

Looking back on scientific journeys can be instructive to others who are
overwhelmed at the prospect of looking forward at their careers (1--5).
By no means is my personal journey over, but since 2002 I have been on a
journey that I did not realize I was on. Now that the paper introducing
the mothur software package is ten years old and has become the most
cited paper published by \emph{Applied and Environmental Microbiology}
(7), it is worth stepping back and using the continued development of
mothur as a story that has parallels to many other research stories.

I fondly recall preparing a poster for the 2002 meeting of research
groups supported by the NSF-supported Microbial Observatories Program. I
wanted to triumphantly show that I had sequenced more than 600 16S rRNA
gene sequences from a single 0.5-g sample of Alaskan soil. This was
greater sequencing depth than anyone else had achieved for a single
sample. As I was preparing the poster, I walked into the office of Jo
Handelsman, my postdoctoral research advisor, and laid out the outline
for the poster. She asked if I could add one of those ``curvy things'',
a rarefaction curve, to show where I was in sampling the community.
Rarefaction curves and attempts to estimate the taxonomic richness of
soil had become popular because of the impactful review by Jennifer
Hughes and her colleagues (8). Their seminal paper introduced the field
to operational taxonomic units (OTUs), rarefaction curves, and richness
estimates. I do not recall whether my poster had a rarefaction curve on
it, but Jo's question primed my career.

\textbf{\emph{Introducing DOTUR and friends.}} When Jo asked me to
generate a rarefaction curve for the poster, the request was not
trivial. How would I bin the sequences into OTUs? Hughes and her
colleagues did it manually and with fewer than 300 sequences. Although I
could possibly do that for my 600 sequences, my goal was to generate
1,000 sequences from the sample and to repeat that sampling effort for
other samples. I needed something that could be automated. Furthermore,
the software that Hughes used, EstimateS (4), required a series of
tedious data formatting steps to perform the analyses we were interested
in performing. I had found my first problem. How would I assign
sequences to OTUs and use that data to estimate the richness and
diversity of a sample? The second problem would involve comparing the
abundance of OTUs found in one sample to another sample. The solution to
the first problem, DOTUR (Distance-based OTUs and Richness), took us two
years to develop (9). DOTUR did two things: given a matrix quantifying
the genetic distance between pairs of sequences, it would cluster those
sequences into OTUs for any distance threshold to define the OTUs and
then it would use the frequency of each OTU to calculate a variety of
alpha diversity metrics. The solutions to the second problem would come
from our work to develop software including \(\int\)-LIBSHUFF (10), SONS
(Shared OTUs and Similarity) (11), and TreeClimber (12). Around the same
time, Catherine Lozupone and Rob Knight were developing their UniFrac
tools to compare communities with a phylogenetic rather than OTU-based
approach (13, 14). With these tools, the field of microbial ecology had
a quantitative toolbox for describing and comparing microbial
communities. Along the way Jo and I would demonstrate the utility of
such tools to answer questions like how many OTUs were there in that
sample of Alaskan soil and how many sequences were needed to sample each
of those OTUs (15)? Where were we in the global bacterial census (16)?
How does the word usage of \emph{Goodnight, Moon} compare to that of
\emph{Portrait of a Lady} and more importantly how is this relevant to
microbial ecology (17)? Most edifying were the more than 2,400 papers
that used DOTUR, SONS, TreeClimber, or \(\int\)-LIBSHUFF to facilitate
their own research questions (Web of Science, 10/1/2019). Had we waited
to solve all of the problems that plagued 16S rRNA gene sequencing, we
would still be waiting.

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification (18--23). We knew there were chimeras (24--27). We knew
that bacteria varied in their \emph{rrn} copy number. Generating a
distance matrix was a prerequisite to using my tools. This wasn't
trivial, but by cobbling together other tools it was possible. We would
assemble, trim and and correct Sanger sequence reads using Chromas or
STADEN (28), align the sequences using ClustalW (29) or ARB (30), check
for chimeras using partial treeing or Bellerephon (27), and calculate a
pairwise distance matrix using DNADIST from the PHYLIP package (31). At
the time, we knew that we only had a loose concept of a species based on
these distances (32). We hoped that an OTU defined as a group of
sequences more than 97\% similar to each other would be a biologically
meaningful unit regardless of whether it fit our notion of a bacterial
species. At the time, I felt that the biggest problems that I could
solve were how to cluster the sequences into OTUs and how to use those
clusterings to test our hypotheses. The only tool available at the time
that automated the clustering step was FastGroup, which implemented a
greedy version of the single linkage algorithm (33). The high cost of
sequencing was also an impediment to experimentation and analysis in
microbial ecology. It was rare for a study design to have experimental
replicates so that one could perform a statistical test to compare
treatment groups. For example, in our testing we frequently used a
dataset comparing Scottish soils from from Alison McCaig and colleagues
(34). This dataset consisted of two experimental groups, each replicated
three times with 45 sequences per replicate. Although great focus has
been placed on the depth of sampling afforded by 454 and Illumina
sequencing, the true benefit of the modern sequencing platforms is the
ability to affordably sequence a large number of technical and
biological replicates. In spite of the many technical challenges, we had
excuses and heuristics to solve problems that served our needs. It is
telling that a recent review of ``best practices'' in generating and
analyzing 16S rRNA gene sequences shows that we still have not solved
many of these issues and that we have even identified additional
problems (35).

As we developed these tools, I found a unique niche in microbiology. My
undergraduate and graduate training as a biological engineer prepared me
to think about research questions from a systems perspective, to think
quantitatively, and to understand the value of using computer programs
to help solve problems. As an undergraduate student, I learned the
Pascal programming language and promptly forgot much of it as a graduate
student and learned MatLab in its place. As a postdoc, I learned the
Perl programming language to better understand how LIBSHUFF, a tool for
comparing the structure of two communities, worked since it was written
in Perl (36). After writing my own version of LIBSHUFF,
\(\int\)-LIBSHUFF and seeing the speed of the version written in C++ by
my collaborator, Bret Larget, I converted my Perl version of DOTUR into
C++. At the time, the conversion from Perl to C++ seemed like an
academic exercise to learn a new language. My Perl version only took a
minute or so to process the final collection of 1,000 sequences and the
C++ version took seconds. Was that really such a big difference? In
hindsight, as we now process datasets with millions of sequences, the
decision to learn to C++ was critical. The ability to pick up computer
languages to solve problems was enabled by my prior training. It was
also a skill that was virtually unheard of in microbiology. Today,
researchers without the ability to program are at a significant
disadvantage (37).

\textbf{\emph{Introducing mothur.}} Shortly after DOTUR was published, I
received an email from Mitch Sogin, a scientist at the Marine Biology
Laboratory (Woods Hole, MA), who asked whether DOTUR could handle more
than a million sequences. Without answering his question, I asked where
he found a million sequences. Little did I know that his email would
represent another pivot in the development of these tools. His group
would be the first to use 454 sequencing technology to generate 16S rRNA
gene sequences (38). Although DOTUR could assign those sequences to
OTUs, at the scale of millions of sequences, it was slow and required a
significant amount of RAM. As I left my postdoc to start my independent
career across the state from Sogin's lab at the University of
Massachusetts in Amherst, my plan was to rewrite DOTUR, SONS,
\(\int\)-LIBSHUFF, and TreeClimber for the new world of massively
parallelized sequencing. The new tool would become mothur.

Milling about at a poster session at an ASM General Meeting in New
Orleans, I ran into Mitch who asked what my plans were for my new lab. I
told him that I wanted to make a tool like ARB, a powerful database tool
and phylogenetics package (30), but for microbial ecology analysis. His
retort was, ``You and what army?'' Up to that point, I had written every
line of code and been answering many emails from people asking for help.
He was right, I would need an army. It would be difficult, but I needed
to learn to let go and share the development process with someone else.
My ``army'' ended up being Sarah Westcott who has worked on the mothur
project from its inception. Today, mothur is over 200,000 lines of code
and Sarah has touched or written nearly every line of it. Beyond writing
and testing mothur's code base, she has become a conduit for many who
are trying to learn the tools of microbial ecology. She patiently
answers questions via email and on the package's discussion forum
(\url{https://forum.mothur.org}). The community and I are lucky that
Sarah has stayed with the project for more than a decade. To be honest,
such dependency on a single person makes the project brittle. In
hindsight, it would have been better to have developed mothur with more
of an ``army'' or team so that there is overlap in people's
understanding of how mothur works. Although a distributed team approach
might work in a software engineering firm, it is not practical in most
academic environments where there is limited funding. There are
certainly projects that make this work, but they are rare.

\textbf{\emph{Competition has been good and healthy.}} mothur has not
been developed in a vacuum and it does not have a monopoly within the
field. As indicated above, each of our decisions were made in the
historical context of the field and with constant pressure from others
developing their own tools for analyzing 16S rRNA gene sequence data.
Competition has been good for mothur and for the field.

From the beginning there have been online tools available at the
Ribosomal Database Project (RDP) (39), greengenes (40), and SILVA (41).
These allowed users a straightforward method of comparing their data to
those collected in a database. There are two primary downsides to these
tools. First, researchers running the online tool must pay the
computational expenses leading to slow process times when hardware
becomes outdated and when numerous users simultaneously attempt their
analysis. Eventually this limitation would result in the termination of
the greengenes website. Second, these platforms provide a
one-size-fits-all analysis. These tools only allow a user to analyze 16S
and in some cases 18S rRNA gene sequences. If a user sequences a
different gene, then the tool will not serve them. These observations
resulted in two design goals we have had with mothur: bringing the
analysis to a user's computer and separating a tool from a specific
database. For example, we commonly use a sequence alignment method that
was originally developed for greengenes (42), but use a SILVA-based
reference alignment because its superior quality (43, 44). In addition,
we offer the \text{na\"ive} Bayesian classifier developed by the RDP
(45) and allow users to train it to any database they want, including
customized databases. In both examples, users can align or classify
non-rRNA gene sequence data. As the bioinformatics tools have matured,
both the RDP and SILVA offer integrated pipelines for analyzing large
datasets, albeit in one-size-fits-all black box implementations.

With the growth in popularity of 16S rRNA gene sequencing there has
naturally been an expansion in the number of people developing tools to
analyze these data. Months after the paper describing mothur was
published, the paper describing QIIME was published (46). Over the past
10 years, many have attempted to create analogies comparing the two
programs: Pepsi vs Coke, Apple vs Windows, etc. It is never clear which
software is which brand and whether the comparisons are meant as a
complement or an insult. Regardless, both programs are very popular.
From my perspective, most of the differences are cosmetic. QIIME is
effectively a bundle of scripts to run other developers' software. For
example, with QIIME (through version 1.9.1), it was even possible to run
mothur through QIIME. One can also run the \text{na\"ive} Bayesian
classifier through QIIME using the original code developed by the RDP.
This caused great frustration for many users because there were numerous
software dependencies that had to be installed. Although the QIIME
developers would go on to create virtual machines and use packaging
tools to simplify installation, these fixes required sophistication by
users who we knew struggled with the basics of navigating a command
line. In contrast, when a user runs mothur, they are running mothur. The
\text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
version of the original code. When we rewrite someone's software we do
it with an eye to improving performance, access, and utility for non-16S
rRNA gene sequence data. For example, while 454 data was popular,
PyroNoise was an effective tool for denoising flowgram data (47).
Running the original code required a large Linux computer cluster and
knowledge of bash and Perl scripting. When we rewrote the code for
mothur, we made it accessible to people using any operating system with
a simple command interface (i.e.~trim.flows and shhh.flows). Our
approach requires significant developer effort, but saves considerable
user effort. As this benefit is multiplied across thousands of projects,
the savings to users has been considerable.

Beyond the large packages like mothur and QIIME, there has been
significant growth in stand alone software tools for sequence curation
(e.g.~PyroNoise (47), PANDAseq (48), DADA2 (49)), chimera checking
(e.g.~UCHIME (50), ChimeraSlayer (51), Perseus (52)), and clustering
(e.g.~USEARCH (53), VSEARCH (54), Swarm (55)). Where possible and when
warrented, we have implemented many of these algorithms directly into
mothur. We have also used this diversity of methods to perform
head-to-head comparisons. Most notable is the area of clustering
algorithms where there have been a large number of algorithms developed
without an obvious method to objectively compare them (56--59). We
applied an objective metric, the Matthew's Correlation Coefficient
(MCC), to evaluate numerous algorithms for clustering sequences into
OTUs. By performing this type of analysis, we were able to objectively
compare the algorithms, make recommendations to the field, and develop
new algorithms that outperformed the existing algorithms. Beyond
evaluating clustering algorithms, we have also evaluated methods of
denoising sequence data (60--62), assessed reference alignments (43,
44), considered the importance of incorporating secondary structure
information in alignments (63), quantified the variation along the 16S
rRNA gene (44), and compared the statistical hypotheses tested by
commonly used tools (64). We have embraced the competition and diversity
of all methods being used to analyze amplicon data as an opportunity to
identify the strengths and weaknesses of the methods in an attempt to
make recommendations to other researchers.

\textbf{\emph{mothur's core principles.}} As mothur has evolved with the
needs of the community, several core principles have emerged that direct
its development. First, mothur is a free, open source software package.
This has been critical in shaping the direction of mothur. We were
content for mothur to be an improved combination of DOTUR and SONS and
leverage existing tools for other steps. Yet, when we learned that the
code for NAST, the algorithm behind the greengenes's aligner (42), was
not open source or publicly available. Similarly, although the SINA
aligner was available through ARB and the SILVA website performed well
it was closed source. Because the ARB implementation did not scale to
large datasets, researchers were left to pay a processing fee to the
SILVA website to align sequences. Thus, we realized that such an
important tool needed to be opened to the community (43). More recently,
the rejection of closed source, commercial tools such as USEARCH can be
seen by the broader adoption of its open source, free competitor,
VSEARCH, within the microbial ecology community (53, 54). Related to
insuring that mothur's code is open source, our second core principle is
that we maintain transparency to our users. Perhaps a user does not need
to interrogate every line of code, but they need to understand what is
happening. Many programs including online workflows encapsulate large
elements of a pipeline in a single command. In contrast, mothur forces
the user to specify each step of the pipeline. Although the former
approach makes an analysis easier for a beginner, it potentially stifles
users that need greater control or understanding of the assumptions at
each step. This control over the pipeline has made it easier for
researchers to customize databases or adapt the pipeline to analyze
non-16S rRNA gene sequence data. Third, as I mentioned above, there has
been a plethora of methods proposed for generating amplicon sequence
data, and curating, aligning, checking for chimeras, classifying, and
clustering the data. I am proud of the data-driven approach we have
taken to comparing these methods. A description of a new method is of
limited value if it is not benchmarked against other methods or control
datasets. Through this core principle and mothur's large reach into the
community, we have helped to develop standards in the analysis of 16S
rRNA gene sequence data. Fourth, a focus on enabling reproducibility has
always been central to the functionality of mothur. From the beginning,
mothur's logfiles have represented a transcript of the user's command
and outputs. When researchers were reluctant to submit sequence data to
the Sequence Read Archive (SRA), we worked with the SRA developers to
create a mothur command (make.sra) to create the package to submit
sequence data through a special mothur portal. A more ambitious project
had its seed on April 1, 2013 when we announced a new ``function'' in
mothur: write.paper. The new command required that the user provide a
454 sff file and a journal title or impact factor. With this
information, mothur would generate a manuscript. This April Fools' Day
joke was poking fun at software that provided an analysis black box but
also at many users' sentiments that data analysis should be so cut and
dry. A few years later, we revisited this concept in the scope of
reproducibility. Why not explicitly script an analysis from downloading
data from the SRA through the rendering of a manuscript ready for
submission? This idea gave rise to the development of the Riffomonas
reproducible research tutorial series that enables researchers to write
their own version of write.paper (65). Perhaps the most important core
principle is that my research group uses mothur to analyze the data we
generate. We ``eat our own dogfood''. This has proven critical as it
again represents transparency and hopefully provides confidence to
mothur's users that we are not making recommendations that we do not
follow ourselves.

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to third parties
to contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists with skills programing in a compiled language
like C++, expecting that community to provide contributors that can
write code in a syntax that prizes execution efficiency over developer
efficiency was not likely. In contrast, the QIIME development team could
be more distributed because their code base was primarily written in
Python, which prizes developer efficiency over execution efficiency.
QIIME is a series of wrappers that allow users to execute other
developers' code making the use of a scripting language like Python
attractive. Their choices resulted in many tradeoffs that have impacted
ease of installation, usability, execution speed, and flexibility. If we
were offered funding to rewrite mothur, we would likely rewrite it as an
R package that leaned heavily on the R language's C++ interface
packages. Of course, such choices are always best in hindsight. Yet,
when we started developing mothur, the ability to interface between
scripting languages like R and Python and C++ code was not as well
developed as it is today. For example, the modern version of the Rcpp
package was first released in 2009 and its popularity was not immediate
(66). Again, the development of mothur has been a product of the
environment that it was created in. Although these decisions have
largely had positive outcomes, there have been tradeoffs that caused us
to sacrifice other goals.

Beyond contributing to the mothur code base, we sought out other ways to
include the community as developers. The paper describing mothur
included 15 co-authors, all but three (Schloss, Westcott, and Ryabin)
responded to a call to provide a wiki page that described how they used
an early version of mothur to analyze a data set. Our vision was that
authors might use the mothur wiki to document reproducible workflows for
papers using mothur but to also provide instructional materials for
other seeking to adapt mothur for their uses
(\url{https://www.mothur.org/wiki}). Unfortunately, once the incentive
of co-authorship was removed, researchers stopped contributing their
workflows to the wiki. Again, this vision and the lack of the
community's adoption of wikis as a mechanism for reporting workflows was
a product of the environment. Although wikis were popular in the late
2000's, they lacked the ability to directly execute the commands that
researchers reported. Such technology would not be possible until the
creation of IPython notebooks (2011) and R markdown (2012). Another
problem with the wiki approach was that potential contributors did not
see the wiki as a community resource. I frequently received emails from
scientists telling me that there was a typo on a specific page when the
intention was that they could correct the typos without my input. We
have been more successful in soliciting input and contributions from the
user community through the mothur discussion forum and GitHub-based
issue tracker. As mothur has matured, we have been dependent on the user
community to use these resources to tell us what features they would
like to see included in mothur and where the documentation is confusing
or incomplete (\url{https://forum.mothur.org}). Often we can count on
people not directly affiliated with mothur to provide instruction and
their own experience to other users on the forum. We are constantly
trying to recruit our ``army'' and are happy to take any contributions
we can. Whether the contributions are to the code base, discussion
forum, or suggestions for new tools, these contributions have been
invaluable to the growth and popularity of mothur.

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and Venn diagrams depicting the
overlap between microbial communities. Such visuals were helpful for
exploring or data; however, I quickly realized that I would never put a
mothur-generated figure into a manuscript I wrote. Such visuals require
far too much customization to be publication-quality. Although QIIME has
incorporated visualization tools through the Emperor package (67), the
challenge of users taking default values has downsides as ordinations
with black background or publishing 3-D ordinations in a 2-D medium
litter the literature. Instead, we have encouraged users to use R
packages to visualize mothur-generated results using the minimalR
instructional materials that I have developed
(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
creation of a graphical user interface (GUI) for running mothur. Forcing
users to interact with mothur through the command line has been a
significant hurdle for many. Unfortunately, the development effort
required to create and maintain a GUI is significant and there is
limited funding for such efforts. The newest version of QIIME (starting
with version 2.0.0) has emphasized interaction with the tools through a
GUI (68) and the related QIITA project offers a web-based GUI (69). It
remains to be seen how this experiment will go. Another downside of
using a GUI is that there is a risk that reproducibility will suffer if
users do not have a mechanism to document their mouse clicks.
Documentation of commands and parameter values is explicit in mothur as
users can provide the software a file with a list of commands and all
commands and output are recorded in a logfile. Given the heightened
focus on reproducibility in recent years (70), we have extended
significant effort in developing instructional materials teaching users
how to organize, document, and execute reproducible pipelines that allow
a user to go from raw sequence data to a compiled manuscript with
figures through the Riffomonas project (65). A final example of a failed
experiment was a collaboration with programmers through Google Summer of
Code to develop commands in mothur that ran the random forest and SVM
machine learning algorithms. Similar to the challenges of developing
attractive visuals, fitting the algorithms' hyperparameters, testing,
and deploying the resulting models require a significant amount of
customization. Furthermore, machine learning is an active area of
research where methods are still being developed and improved.
Thankfully, there are numerous R and Python packages that do a better
job of developing these models (71, 72). Again, we have put our efforts
into developing instructional materials that mothur users can use to fit
such models to their data. In each of our ``failed'' experiments, the
real problems were straying from what mothur does well and failing to
grasp what we really wanted the innovation to do.
